rang: Reconstructing reproducible R computational environments

### 2023-05-18 paper/paper.qmd

[^r]: In this paper, we will focus on R, a popular programming language used frequently in various computational fields (e.g. computational social science, bioinformatics).

In reality, the impact of Component A is relatively weak as mainstream, open source programming languages and their software libraries are usually cross platform. In modern computational research, Linux is the de-facto operating system in High Performance Computing environments. Instead, the impact of Components B, C, and D is much higher. Component D is the most volatile among them all as there are many possible combinations of R packages and versions. Software updates with breaking changes (even in a dependency) might render existing shared code using those changed features not executable or not producing the same result anymore. Also, software obsolescence is commonplace, especially since academic software is often not well maintained due to lack of incentives [@merow:2023:B].

@@

There has been many papers written on how containerization solutions such as Docker can be helpful also to foster computational reproducibility of science [e.g. @nuest:2019;@peikert:2021:RDA;@boettiger:2017:IR]. Although tutorials are available [e.g. @nuest:2019], providing a declarative description of the computational environment in the form of `Dockerfile` is far from the standard code sharing practice. This might be due to a lack of (DevOps) skills of most scientists to create a `Dockerfile` [@kim:2018:E]. But there are many tools available to automate the process [e.g. @nuest:2019]. The case in point described in this paper, `rang`, is one of them. We argue that `rang` is the only easy-to-use solution available that can specify and restore all four components without the reliance on any commercial service <!-- such as MRAN -->.

@@

It is also worth mentioning that MRAN is not the only archival service. Posit also provides a free (*gratis*) time-stamped daily backup of CRAN and Bioconductor (a series of repositories of R package for bioinformatics and computational biology) called Posit Public Package Manager [^posit]. It can goes as far back as October 10, 2017.

@@

These solutions are better for prospective usage, i.e. using them now to ensure the reproducibility of the current research for future researchers. `rang` mostly targets retrospective usage, i.e. using `rang` to reconstruct historical R computational environments for which the declarative descriptions are not available. One can think of `rang` as an archaeological tool. In this realm, we could not find any existing solution targeting R specifically which does not currently depend on MRAN.

@@

For R >= 3.1, the base images from the Rocker project are used [@boettiger:2017:IR]. For R < 3.1 but >= 1.3.1, a custom base image based on Debian is used. As of writing, `rang` does not support R < 1.3.1, i.e. snapshot date earlier than 2001-08-31 (which is 13 years earlier than all solutions depending on MRAN). There are two features of `dockerize()` that are important for future reproducibility.

1. By default, the Docker image building process downloads source packages from their sources and then compiles them. This step depends on the future availability of R packages on CRAN (which is extremely likely to be the case in the near future, given the continuous availability since 1997-04-23) [^1], Bioconductor, and Github. However, it is also possible to cache (or archive) the source packages now. The archived R packages can then be used instead during the building process. The significance of this step in terms of long-term computational reproducibility will be discussed in Section 4.

@@

When `as_pkgrefs()` is applied to a single path of a directory, it scans all relevant files (`DESCRIPTION`, R scripts and R Markdown files) for all R packages used (based on `renv::dependencies()` <!-- [^gh] -->). How it works is demonstrated in three of the following examples below. But an important caveat is that it can only scan CRAN and Bioconductor packages.

@@

The following are some examples of how `rang` can be used to make shared, but otherwise unexecutable, R code runnable again. The examples were drawn from various fields spanning from political science, psychological science, and bioinformatics.

@@

The issue is that installing `quanteda` 1.3.4 this way installs the latest dependencies from CRAN. `quanteda` 1.3.4 uses a deprecated (but not yet removed) function of `Matrix` (`as(<dgTMatrix>, "dgCMatrix")`). If this function were removed in the future, the above code snippet would not be executable anymore.

@@

Inside the container, the materials are located in the `materials` directory. We used the following shell script to test the reproducibility of all R scripts.

@@

Using `rang`, we are able to reconstruct a computational environment with R 2.15.0 (2012-03-30) to run all code snippets published in Jurka [@jurka:2012] [^speed]. For removed CRAN packages, we strongly recommend querying the Github read-only mirror of CRAN instead (https://github.com/cran). It is because in this way, the resolved system requirements have a higher chance of being correct.

@@

The software paper of the R package `ptproc` was published in 2003 and introduced multidimensional point process models [@peng:2003:MDP]. But the package has been removed from CRAN for over a decade (at least). The only release on CRAN was on 2002-10-10. The package is still listed in the "Handling and Analyzing Spatio-Temporal Data" CRAN Task View [^taskview] despite being uninstallable without modification on any modern R system (see below). As of writing, the package, as a tarball file (tar.gz), is still downloadable from the original author's website [^peng].

@@

Even with this over-a-decade removal and new packages with similar functionalities have been created, there is evidence that `ptproc` is still being sought for. As late as 2017, there are blog posts on how to install the long obsolete package on modern versions of R [^blog]. The package is extremely challenging to install on a modern R system because the package was written before the introduction of name space management in R 1.7.0 [@RN-2003-001]. In other words, the available tarball files from the original author's website and CRAN do not contain a `NAMESPACE` file like all other modern R packages do.

@@

The preparation of the research compendium is easy as `rang` can scan a materials directory for all R packages used [^dmetar].

@@

This paper presents `rang`, a solution to (re)construct R computational environments based on Docker. As the six examples in Section 3 show, `rang` can be used archaeologically to rerun old code, many of them not executable without the analytic and reconstruction processes facilitated by `rang`. These retrospective use cases demonstrate how versatile `rang` is. `rang` is also helpful for prospective usage, as demonstrated in Section 4 whereby an executable compendium is created.

There are still many features that we did not mention in this paper (see Table 1). `rang` is built with interoperability in mind. As of writing, `rang` is interoperable with existing R packages such as `renv` and R built-in `sessionInfo()`. Also, the `rang` object can be used for network analysis with R packages such as `igraph`.

Computational reproducibility is a complex topic and as in all of these complex topics, there is no silver bullet [@canon:2019:CPR]. All solutions have their trade-offs and cannot solve all issues. Specifically, `rang` can restore a specific computational environment but there might be potential barriers to reproducibility related to external dependencies that cannot be containerized. A common external dependency is an external data source connected via a web application programming interface (API). Some examples are the so-called "API packages" such as `STRINGdb` (an R interface to the STRING protein-protein database), `rtoot` (an R interface to the Mastodon API), and `googleLanguageR` (an R interface to the Google Translation API). Although these API packages can be containerized, the external data sources cannot. Therefore, `rang` is not a solution to issues related to external dependencies. For data collection via these API packages, we recommend caching the collected data and sharing the cached data. For computation via API packages (e.g. Google Translate), we recommend using alternative methods that do not require external dependencies (e.g. [@chan:2020:REC]).

>>>>

[^r]: In this paper, we will focus on R, a popular programming language used frequently in various computational fields (e.g. computational social science or bioinformatics).

In reality, the impact of Component A is relatively weak as mainstream open source programming languages and their software libraries are usually cross platform. In modern computational research, Linux is the de-facto operating system in High Performance Computing environments. Instead, the impact of Components B, C, and D is much higher. Component D is the most volatile among them all as there are many possible combinations of R packages and versions. Software updates with breaking changes (even in a dependency) might render existing shared code using those changed features not executable or not producing the same result anymore. Also, software obsolescence is commonplace, especially since academic software is often not well maintained due to lack of incentives [@merow:2023:B].

@@

There has been many papers written on how containerization solutions such as Docker can be helpful also to foster computational reproducibility of science [e.g. @nuest:2019;@peikert:2021:RDA;@boettiger:2017:IR]. Although tutorials are available [e.g. @nuest:2019], providing a declarative description of the computational environment in the form of `Dockerfile` is far from the standard code sharing practice. This might be due to a lack of (DevOps) skills of most scientists to create a `Dockerfile` [@kim:2018:E]. But there are many tools available to automate this process [e.g. @nuest:2019]. The case in point described in this paper, `rang`, is one of them. We argue that `rang` is the only easy-to-use solution available that can specify and restore all four components without the reliance on any commercial service <!-- such as MRAN -->.

@@

It is also worth mentioning that MRAN is not the only archival service. Posit also provides a free (*gratis*) time-stamped daily backup of CRAN and Bioconductor (a series of repositories of R package for bioinformatics and computational biology) called Posit Public Package Manager [^posit]. It can go as far back as October 10, 2017.

@@

These solutions are better for prospective usage, i.e. using them now to ensure the reproducibility of the current research for future researchers. `rang` mostly targets retrospective usage, i.e., to reconstruct historical R computational environments for which the declarative descriptions are not available. One can think of `rang` as an archaeological tool. In this realm, we could not find any existing solution targeting R specifically which does not currently depend on MRAN.

@@

For R >= 3.1, the base images from the Rocker project are used [@boettiger:2017:IR]. For R < 3.1 but >= 1.3.1, a custom base image based on Debian is used. As of writing, `rang` does not support R < 1.3.1, i.e. snapshot dates earlier than 2001-08-31 (which is 13 years earlier than all solutions depending on MRAN). There are two features of `dockerize()` that are important for future reproducibility.

1. By default, the Docker image building process downloads source packages from their sources and then compiles them. This step depends on the future availability of R packages on CRAN (which is extremely likely to be the case in the near future, given the continuous availability since 1997-04-23) [^1], Bioconductor, and Github. However, it is also possible to cache (or archive) the source packages now. The archived R packages can then be used during the building process instead. The significance of this step in terms of long-term computational reproducibility will be discussed in Section 4.

@@

When `as_pkgrefs()` is applied to a single path of a directory, it scans all relevant files (`DESCRIPTION`, R scripts and R Markdown files) for all R packages used (based on `renv::dependencies()` <!-- [^gh] -->). Its functionality is demonstrated in three of the following examples below. However, an important caveat is that it can only scan CRAN and Bioconductor packages.

@@

In this section, we present several examples for how `rang` can be used to make shared, but otherwise unexecutable, R code runnable again. The examples were drawn from various fields spanning from political science, psychological science, and bioinformatics.

@@

The problem is that installing `quanteda` 1.3.4 this way installs the latest dependencies from CRAN and not the "period appropriate" ones. `quanteda` 1.3.4 uses a deprecated (but not yet removed) function of `Matrix` (`as(<dgTMatrix>, "dgCMatrix")`). If this function is removed in the future, the above code snippet will not be executable anymore.

@@

Inside the container, the material is located in the `materials` directory. We used the following shell script to test the reproducibility of all R scripts.

@@

Using `rang`, we are able to reconstruct a computational environment with R 2.15.0 (2012-03-30) to run all code snippets published in Jurka [@jurka:2012] [^speed]. For removed CRAN packages, we strongly recommend querying the Github read-only mirror of CRAN instead (https://github.com/cran). This way, the resolved system requirements have a higher chance of being correct.

@@

The software paper of the R package `ptproc` was published in 2003 and introduced multidimensional point process models [@peng:2003:MDP]. But the package has been removed from CRAN for over a decade (at least). The only release on CRAN was on 2002-10-10. The package is still listed in the "Handling and Analyzing Spatio-Temporal Data" CRAN Task View [^taskview] despite being uninstallable without modification on any modern R system (see below). As of writing the package is still downloadable from the original author's website[^peng] as a tarball file (tar.gz).

@@

Even with this over-a-decade removal and new packages with similar functionalities have been created, there is evidence that `ptproc` is still being sought for. As late as 2017, there are blog posts on how to install the long obsolete package on modern versions of R [^blog]. The package is extremely challenging to install on a modern R system because it was written before the introduction of name space management in R 1.7.0 [@RN-2003-001]. In other words, the available tarball files from the original author's website and CRAN do not contain a `NAMESPACE` file like all other modern R packages do.

@@

The preparation of the research compendium is easy since `rang` can scan a materials directory for all R packages used [^dmetar].

@@

This paper presents `rang`, a solution to (re)construct R computational environments based on Docker. As the six examples in Section 3 show, `rang` can be used archaeologically to rerun old code, many of them not executable without the analytic and reconstruction processes facilitated by `rang`. These retrospective use cases demonstrate how versatile `rang` is. `rang` is also helpful for prospective usage, as demonstrated in Section 4 where an executable compendium is introduced.

There are still many features that we did not cover in this paper (see Table 1). `rang` is built with interoperability in mind. As of writing, `rang` is interoperable with existing R packages such as `renv` and R built-in `sessionInfo()`. Also, the `rang` object can be used for network analysis with R packages such as `igraph`.

Computational reproducibility is a complex topic and as in all of these complex topics, there is no silver bullet [@canon:2019:CPR]. All solutions have their trade-offs and cannot solve all issues. Although `rang` can restore a specific computational environment,  there might be potential barriers to reproducibility related to external dependencies which cannot be containerized. A common external dependency is a data source connected via a web application programming interface (API). Some examples are so-called "API packages" such as `STRINGdb` (an R interface to the STRING protein-protein database), `rtoot` (an R interface to the Mastodon API), and `googleLanguageR` (an R interface to the Google Translation API). Although these API packages can be containerized, the external data sources cannot. Therefore, `rang` is not a solution to issues related to external dependencies. For data collection via these API packages, we recommend caching the collected data and sharing the cached data. For computation via API packages (e.g. Google Translate), we recommend using alternative methods that do not require external dependencies (e.g. [@chan:2020:REC]).
